Collaborative Visual Analysis on the Web with RCloud

### 2015-06-25 discussion.tex
working

\paragraph*{Shared artifacts and artifact histories} are a central feature
of RCloud, through shared workbooks. We observed that hackers readily
share experiments, demonstrate techniques, publish results to peers and
managers, and transfer algorithms to other groups. Artifact histories
can be accessed through the notebook tree or through GitHub's web interface.
But a problem is that it is easy to create a large number of artifacts
and the system does not help enough to organize and navigate them.

\paragraph*{Modularity and granularity} Modularity, or the ability
to partition work into independent units, is a key to working
productively in teams.
It is a benefit if the units can be kept small or granular, so team members
can realize benefits at least proportional to the work on the units.
RCloud's notebook and versioning capabilities allow work to be divided
into units as fine as the underlying language allows, and versioning
encourages making incremental changes at low cost and without disrupting
the work of others.

\paragraph*{View sharing, bookmarking} Most resources in RCloud are named
and accessed as URLs. This proved to be effective for sharing analyses
and integrating them with external processes.  It is particularly
advantageous for work to be shared as URLs that provide access to code
and experiments, instead of by pasting screenshots into reports.

@@

\paragraph*{Content export} is not a capability we aimed at supporting,
and R already has many packages for this. Recently, due to popular
demand, we added user interface support for exporting plot images.
We assume that RCloud notebooks should play well in the ecosystem
of other tools, but sometimes it is difficult to know whether adding
a feature for compatibility will offload complexity, or increase it.

\paragraph*{Social-psychological incentives} and {\bf voting and ranking} are
supported through starring and forking. These mechanisms are employed often
on the platform. An obvious next step would be to enhance recommendations
using relationships discovered by static and dynamic code analysis. This
may be considered both within and across collections of scripts (the latter
being similar to VisTrails' enhanced recommendations by clustering multiple
workflows).  It seems valuable to know which packages are frequently used
together, or appear in proximity to a certain record types or data feeds.
In general, trivial or passive mechanisms to collect data for
recommendations are essential.

\paragraph*{Group management, size and diversity} This area needs better
support in RCloud, but is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. RCloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

\paragraph*{Curation} Even without formal group management, users may
organize groups of related notebooks using tree folders.
We found this particularly effective in collecting and distributing
training materials.

@@

\paragraph*{Versioning and Upgrading}
Some aspects of the programming environment are more difficult to
manage in RCloud than in conventional systems. RCloud does not separate
the development and deployment environments, and every version of every
prototype is shared by default. While this encourages
collaboration, it also quickly exposes errors such as
misconfiguration, mismatches between package versions, and
programming errors, that can unintentionally affect production
websites. For example, if an RCloud website is distributed across
multiple servers, conflicting versions of system software
can cause errors that are hard to debug.

Versioning of software components must be managed in several
levels: in data analysis and visualization scripts (the
notebooks themselves); in the R environment such as the installed
libraries and packages; and in the external environment such as the
operating system, libraries, and protocols spoken by remote
services. At one end of the scale, we have full control over notebooks
via git, along with conventions to fix stable notebook versions.
The R environment is under the control of its package manager that
has rules to ensure reasonable consistency.
(It is possible for RCloud to access information about package
versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not
strong.) At the other end of the scale, there is hardly any version
control for the external environment, and if an external service 
used by a notebook changes its API, the notebook can stop working.

@@

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls; web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.
%
Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control for information security is delegated to the host
operating system and web server.
Most operating systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome. This means that information security
in RCloud is a delicate, unsolved matter. More
sophisticated approaches, such as information flow analysis,
and query languages that ensure differential privacy, are
open research topics~\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

@@

Surprisingly, we found that when users realize that \emph{scratch work
is getting saved}, the extra versions become burdensome to them.
Clearly, RCloud is an extreme, and the right solution is
in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. 
Bookkeeping would be reduced to selecting the \emph{data} to be kept,
and the code that generated it would be derived and saved transparently.

>>>>

\paragraph*{Shared artifacts and artifact histories} are a central feature
of RCloud, through shared workbooks. We observed that hackers readily
share experiments, demonstrate techniques, publish results to peers and
managers, and transfer algorithms to other groups. Artifact histories
can be accessed through the notebook tree or through GitHub's web interface.
But it is easy to create a large number of artifacts
and the system does not help enough to organize and navigate them.

\paragraph*{Modularity and granularity} Modularity, or the ability
to partition work into independent units, is a key to working
productively in teams.
It is a benefit if the units can be kept small or granular, so team members
can realize benefits at least proportional to the work on the units.
RCloud's notebook and versioning allow work to be divided
into units as fine as the underlying language allows, and
encourage making incremental changes at low cost and without disrupting
the work of others.

\paragraph*{View sharing, bookmarking} Most resources in RCloud are named
and accessed as URLs. This proved to be effective for sharing analyses
and integrating them with external processes.  It is particularly
advantageous for work to be shared as URLs that provide access to live
notebooks, instead of by pasting screenshots into reports.

@@

\paragraph*{Content export} is not a capability we aimed at supporting,
and R already has many packages for this. Recently, due to popular
demand, we added user interface support for exporting plot images.
RCloud notebooks should play well in the R ecosystem,
but sometimes it is difficult to know whether adding
a feature for compatibility will offload complexity, or increase it.

\paragraph*{Social-psychological incentives} and {\bf voting and ranking} are
supported through starring and forking. These mechanisms are employed often
on the platform. An obvious next step would be to enhance recommendations
using relationships discovered by static and dynamic code analysis. This
may be considered both within and across collections of scripts (the latter
being similar to VisTrails' enhanced recommendations by clustering multiple
workflows). It seems helpful to help users find which packages are frequently
used together, or when using record types or data feeds.
Overall, simple, passive mechanisms to collect data for recommendations are preferred.

\paragraph*{Group management, size and diversity} This area needs better
support, and is clearly important to working in teams.
We rely on external administrative processes and social conventions
to manage accounts and groups. RCloud could benefit greatly from
integrating social media to track identities and groups and to maintain
communication channels, instead of having its own isolated solution.

\paragraph*{Curation} Even without formal group management, users often
organize groups of related notebooks using tree folders.
We found this particularly effective in collecting and distributing
training materials.

@@

\paragraph*{Versioning and Upgrading}
Some aspects of the environment are more difficult to
manage in RCloud than in conventional systems. RCloud does not separate
its development and deployment environments, and every version of every
prototype is shared by default. Although this encourages
collaboration, it also quickly exposes misconfigurations,
mismatches between package versions, and programming errors
that can unintentionally affect production websites.
If an RCloud website is distributed across multiple servers,
conflicting versions of system software
can also cause errors that are hard to debug.

Versioning of software components must be managed in several
levels: in data analysis and visualization scripts (the
notebooks themselves); in the R environment such as the installed
libraries and packages; and in the external environment such as the
operating system, libraries, and protocols spoken by remote
services. At one end of the scale, we have full control over notebooks
via git, along with conventions to fix stable notebook versions.
The R environment is under the control of its package manager that
has rules to ensure reasonable consistency.
% (It is possible for RCloud to access information about package
% versions and configurations, but support for checking compatibility
% and maintaining multiple versions in the same environment is not
% strong.)
At the other end of the scale, there is basically no
control over the external environment, and if an external service 
used by a notebook changes its API, the notebook can stop working.

@@

\paragraph*{Security}
Because RCloud allows arbitrary R code to be run
\emph{by design}, it's not straightforward to protect deployments
from running unauthorized arbitrary code. RCloud uses an object capability model
\cite{Miller:2006:RCT} recently added to the Rserve
protocol~\cite{Urbanek:2003:AFW}. Object capabilities are opaque handles
to remote procedure calls; web browsers never directly instruct the
RCloud backend to execute arbitrary code, preventing unauthenticated clients
from making unauthorized calls to the RCloud runtime environment.
%
Our back-end environment, on the other hand, assumes a high degree of trust between users:
access control is delegated to the host operating system and web server.
Most operating systems rely either on coarse permission models,
which tend to be ineffective, or on detailed access control
lists, which tend to be cumbersome.
Information security in RCloud today is an unsolved matter.
More sophisticated approaches are open research topics~\cite{Moore:2011:SAF}.
% We are hopeful that having a richer
% environment for collaboration and information sharing will encourage
% new approaches to information security in visual analytics.
%\gordon{This sounds like we think we are helping by pushing the
%system to the breaking point, is that the idea?}

@@

% Surprisingly, we found that when users realize that \emph{scratch work
We found that when users realize that \emph{scratch work
is getting saved}, it sheer volume becomes burdensome.
Clearly, RCloud is an extreme, and the right solution is
probably in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. 
Bookkeeping would be reduced to selecting the \emph{data} to be kept,
and the code that generated it would be derived and saved transparently.

### 2015-06-25 interviews.tex
working

Some users, who tried RCloud and were not able to continue for organizational
reasons, miss certain capabilities. Leith ``[likes] the concept of being
able to create notebooks and share them... A wiki is not the best way for
communicating results - it's kind of like writing a blog post with very limited
functionality. I have to save every picture and post it as an image.'' She said,
``I can't share all of the code because it would just get crowded and wouldn't
look right on a wiki.'' Iris explains, ``If I could make a folder on RCloud and
have Python notebooks and also Pig notebooks there, and execute them from RCloud,
that would be much better than my current [environment], because that would free
me from manual documentation and version control and also telling people where
my code was. It would be just, hey, go look on RCloud, here's the stuff.''

@@

The problem of dead notebooks is compounded by changing
packages and example notebooks at the same time. Eric will ``go back and
change [a notebook] and update the page, but in the meantime people
have forked it. I'll have to change a package as well, so their
old fork stops working, and they complain.''

Eric tried to keep his notebook tree well organized, but this didn't help,
because some people kept old links in their email, or forked notebooks which 
had become obsolete. By design, notebooks that are deleted in the user interface
are not purged from the repository. He says he became reluctant to share notebooks:
``Do I want to support this forever?''

>>>>

Some users, who tried RCloud and were not able to continue for organizational
reasons, miss certain capabilities. Leith ``[likes] \comment{the concept of} being
able to create notebooks and share them... A wiki is not the best way for
communicating results - it's \comment{kind of} like writing a blog post with very limited
functionality. I have to save every picture and post it as an image.'' She said,
``I can't share all of the code because it would just get crowded and wouldn't
look right on a wiki.'' Iris explains, ``If I could make a folder on RCloud and
have Python notebooks and also Pig notebooks there, and execute them from RCloud,
that would be much better than my current [environment], because that would free
me from manual documentation and version control and also telling people where
my code was. It would be just, hey, go look on RCloud, here's the stuff.''

@@

The problem of dead notebooks is compounded by changing
packages and example notebooks at the same time. Eric will ``go back and
change [a notebook] and update the page, but in the meantime people
have forked it. I'll have to change a package as well, so their
old fork stops working, and they complain.''
% we don't need a new paragraph here - -scn
Eric tried to keep his notebook tree well organized, but this didn't help,
because some people kept old links in email, or forked notebooks which 
had become obsolete.
% By design, notebooks that are deleted in the user interface
% are not purged from the repository.
He says he became reluctant to share notebooks: ``Do I want to support this forever?''

### 2015-06-25 introduction.tex
working

\item Technology transfer.
In most organizations, development of analyses and visualizations 
s done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
benefit from tools developed by hackers and scripters.
Results of experiments are often shared first by copying screen
shots or static output. By the time decisions are made based on
the results of those screenshots, data and processes may have changed
so much that the decision can be wrong.
\stephen{Transitional sentence here.}
Generally, the connection between these two worlds is made by IT staff,
who port or rewrite code so it can run as a stable production service.
In an environment where business needs can change rapidly, this
process does not scale. It is also difficult to trace problems
from production back to EDA.
Our goal is to merge the worlds of EDA and production to close this gap.

@@

A key point is that, for the most part, the improved capabilities are
available by default, without much explicit involvement on the part of
data scientists and other customers. Visual analytics experiments are
conveniently shared and turned into production websites, without moving
or porting code. All the published artifacts in the system can be
searched immediately. Recommendation is as easy as clicking a star
on a useful notebook.

>>>>

\item Technology transfer.
In most organizations, development of analyses and visualizations 
is done by \emph{hackers} and \emph{scripters} (adopting Kandel et al.'s
terminology~\cite{Kandel:2012:EDA}). \emph{Application users} 
benefit from tools developed by hackers and scripters.
Results of experiments are often first shared by copying screen
shots or static output. By the time decisions are made based on
those screenshots, the data and processes may have
changed so much that the decision can be wrong.
Moreover, the connection between EDA and deployment
are made by porting or rewriting code so it runs as a stable service.
In an environment where business needs can change rapidly, this
process does not scale. It is also difficult to trace problems
from production back to EDA.
Our goal is to merge the worlds of EDA and deployment, to close the gap.

@@

For the most part, the improved capabilities are available by default,
without much explicit involvement on the part of
data scientists and other customers. Visual analytics experiments are
conveniently shared and turned into production websites, without moving
or porting code. All the published artifacts in the system can be
searched immediately. Recommendation is as easy as clicking a star
on a useful notebook.

### 2015-06-25 system.tex
working

One of the main contributions of RCloud is the notion that notebooks are
``always deployed''. In other words, the most recent version of a
notebook is immediately available to all other users of the system.
Another way to describe this is that
RCloud lacks a ``save'' button: any notebook cell that runs is always
associated to a notebook version serialized to disk. Although one of
our interview subjects reported that this sometimes leads to an
excessively fine-grained sequence of versions (see
Section~\ref{sec:interviews}), we believe that the alternative of
losing automatic sharing is worse (and see
Section~\ref{sec:interviews} again for users reporting their
satisfaction over the low-friction shareability).

@@

One side benefit of centralizing the execution and storage of
notebooks is that it becomes feasible to collect usage information
that might be lost in a conventional environment.
In the case of social data visualization
platforms, we would like to exploit usage data to help analysts
find content of interest, whether the content is actual source code
or accompanying data. The standard way to achieve this is through
\emph{user-generated curation} and \emph{automatic recommendations}.

Automatic recommendations have become famous in the user experience
provided by companies such as Amazon and Netflix (``If you liked
that film, then you should like this one too''). To make such
recommendations, we need users to \emph{curate} the collection, by
providing explicit reviews or some other method of indicating interest.
%%
%% In RCloud, reputation and interest are a relationship between
%% \emph{notebooks} and \emph{users}, rather than a relationship between
%% user pairs. We chose this approach because we expect initial 
%% RCloud deployments to have relatively few users, but some users to
%% create many notebooks. Under that assumption, assigning interest
%% to users would not provide sufficiently ``high-resolution'' data.
%%
We incorporate both explicit and implicit indications of interest
in notebooks. Explicit interest is indicated by ``starring,'' or
clicking on a button that marks a notebook as interesting.
This makes explicit indication of interest a nearly trivial operation,
always available and easy to use.
%
RCloud today uses only simple counts of stars to measure overall
notebook interest, mostly because standard
recommender system algorithms 
require reasonably large training sets to pay off.
Nevertheless, the starring mechanism is sufficient
to create personalized recommendations~\cite{Hu:2008:CFF}.

Implicit signaling of interest is supported by keeping click-through
counts~\cite{Joachims:2005:AIC} and execution counts. In addition to
these standard techniques, we hope to apply static and dynamic code
analysis to infer fine-grained information about relationships,
for example, which packages and data sets often appear together,
% in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
in the style of Codex.~\cite{Fast:2014:ECS}.

>>>>

One of the main contributions of RCloud is the notion that notebooks are
``always deployed''. In other words, the most recent version of a
notebook is immediately available to all other users of the system.
Another way to describe this is that
RCloud lacks a ``save'' button: any notebook cell that runs is always
associated to a notebook version serialized to disk. Although one of
our interview subjects reported that this sometimes leads to an
excessively fine-grained sequence of versions (see
Section~\ref{sec:interviews}), 
losing information that could be shared is worse (see
Section~\ref{sec:interviews} user comments on
satisfaction with low-friction shareability).

@@

A side benefit of centralizing the execution and storage of
notebooks is that it becomes feasible to collect usage information
that is lost in conventional environments.
In the case of social data visualization
platforms, we would like to exploit usage data to help analysts
find content of interest, whether that content is source code
or data. Standard ways to achieve this are through
\emph{user-generated curation} and \emph{automatic recommendations}.

Automatic recommendations have become famous in the user experience
provided by companies such as Amazon and Netflix (``If you liked
that film, you'll like this one too''). To make such
recommendations, we need users to \emph{curate} the collection, by
providing explicit reviews or some other method of indicating interest.
%%
%% In RCloud, reputation and interest are a relationship between
%% \emph{notebooks} and \emph{users}, rather than a relationship between
%% user pairs. We chose this approach because we expect initial 
%% RCloud deployments to have relatively few users, but some users to
%% create many notebooks. Under that assumption, assigning interest
%% to users would not provide sufficiently ``high-resolution'' data.
%%
We incorporate both explicit and implicit indications of interest
in notebooks. Explicit interest is indicated by ``starring,'' or
clicking on a button that marks a notebook as interesting.
This makes explicit indication of interest a nearly trivial operation,
always available and easy to use.
%
RCloud today uses only simple counts of stars to measure overall
notebook interest, mostly because standard
recommender system algorithms 
require reasonably large training sets to pay off.
Nevertheless, the starring mechanism is sufficient
to create personalized recommendations~\cite{Hu:2008:CFF}.

Implicit signaling of interest is supported by keeping click-through
counts~\cite{Joachims:2005:AIC} and execution counts. In addition to
these standard techniques, we hope to apply static and dynamic code
analysis to infer fine-grained information about relationships,
for example, which packages and data sets often appear together,
% in the style of Fast et al's Codex system.~\cite{Fast:2014:ECS}.
in the style of Codex~\cite{Fast:2014:ECS}.
